---
title: "Data Wrangling Assignment"
author: "Nicole Baker"
date: "2023-03-23"
output: 
  html_document: default
  word_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Wrangling

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

### I am using the microbiome data:
```{r}
microbiome.data<-read.csv("Microbiome.csv")
microbiome.data
```

## 1. Selecting specific columns of data

```{r}
microbiome.subset<-select(microbiome.data,OTU,
                          SampleID,Abundance,Crop,
                          Compartment,DateSampled,
                          GrowthStage,Treatment,Rep,
                          Fungicide,Kingdom,Phylum,
                          Class,Order,Family,Genus,
                          Species,Taxonomy)
microbiome.subset
```

## 2. Calculate mean % relative abundance by compartment, using pipe

you need to mutate data by making a percent column (multiply Abundance values by 100). Then a summarise statement to calculate the mean percentage.
In order to group this by Compartment, you need to add a group_by statement prior to the mutate statement.

```{r}
microbiome.subset %>%
  group_by(Compartment) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```

now, you are given means for root and leaf compartments.

## 3. Calculate mean % relative abundance by compartment and fungicide, using pipe

one way to do this is by adding both Compartment and Fungicide to the group_by() statement, producing four mean values: 

```{r}
microbiome.subset %>%
  group_by(Compartment,Fungicide) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```
 

another way to do this is by adding Compartment and Fungicide to separate group_by() statements, which produces only two mean values: 

```{r}
microbiome.subset %>%
  group_by(Compartment) %>%
  group_by(Fungicide) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```

## 4. restricting previous analysis to just fungal class "Dothideomycetes"

In order to do this, add the filter function underneath the microbiome.subset statement:
**I am using the first example of grouping, where compartment and fungicide are in the same group_by() statement**

```{r}
microbiome.subset %>%
  filter(Class == "Dothideomycetes") %>%
  group_by(Compartment,Fungicide) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```

## 5. Same as above, grouped by Orders in genus Dothideomycetes

switch the group_by() statement to have Order:

```{r}
microbiome.subset %>%
  group_by(Compartment,Fungicide,Order) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```

Additionally, you could just have a group_by() statement with just Order in it:

```{r}
microbiome.subset %>%
  group_by(Order) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent)) %>% 
  
  head()
```

## 6. Calculate standard error for each mean in question 5

Add in a calculation for standard deviation in the summarise function, and then use it in another muatation calculating a column for standard error. 

**I used the second example where Order is by itself in the group_by() statement.**

```{r}
microbiome.subset %>%
  group_by(Order) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent),
            n=n(),
            stddev=sd(Percent)) %>%
  mutate(stderr=stddev/sqrt(n)) %>%

head()

```

## 7. Select columms, pivot data to wide so that compartment are column headers

to make the data wide formatm use pivot_wider():
**I used the names from Compartment, and the values from Mean, so that the mean values for each order were in separate columns titled "leaf" and "root".**

```{r}
microbiome.subset %>%
  filter(Class=="Dothideomycetes") %>%
  group_by(Order,Compartment,Fungicide) %>%
  mutate(Percent=Abundance*100) %>%
  summarise(Mean=mean(Percent),
            n=n(),
            stddev=sd(Percent)) %>%
  mutate(stderr=stddev/sqrt(n)) %>%
  pivot_wider(names_from=Compartment,values_from=Mean)
```

## 8. recreate the given plot from question 7 dataset

I need to add the ggplot function and layers, and I also need to reorder the y axis.

```{r}
microbiome.subset %>%
  filter(Class=="Dothideomycetes") %>%
  group_by(Order,Compartment,Fungicide) %>%
  mutate(Percent=Abundance*100) %>%
  mutate(Order2=fct_reorder(Order,desc(Percent))) %>%
  ggplot(aes(x=Percent,y=Order2,fill=Fungicide,color=Fungicide))+
  stat_summary(fun=mean,geom="bar",position="dodge")+
  ylab("Order")+
  xlab("Percent relative abundance")
  
  

```


**For some reason, the fct_reorder() part of code will not work for me. the y axis doesn't change whether I put a - before Percent or not. but the graph otherwise looks fine.**
 




